Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-02-21 manuscript/HIAs.qmd
change municipalities

# Some directories
mireFile <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"

path_infrastructure <- ifelse(dir == "C:",
  "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif",
  "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif"
)

path_naturetypes <- ifelse(dir == "C:",
  "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
  "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb"
)

path_muni <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")

@@

@@

# I already did some work to identify the relevant nature types

# summary file (https://github.com/NINAnor/ecosystemCondition/blob/main/data/naturetypes/natureType_summary.rds)
naturetypes_summary <- readRDS("../data/natureType_summary.rds")

@@

```{r natureTypeData}
#| eval: false

# The data is too big to be stored on GitHub

# Impart polygon data set
naturetypes <- sf::st_read(dsn = path_naturetypes)
# 117k polygons (2022)
# sometimes this function struggles with Norwegian letters.
# Restart R and try again.

naturetypes <- naturetypes |>
  # keep only wetlands
  filter(
    hovedøkosystem == "våtmark",
    naturtype %in% nts,
    naturtype != "Kalkrik helofyttsump"
  ) |>
  # calculate the areas (m2) of the polygons
  mutate(area = SHAPE |> st_area()) |>
  # the variable codes and values are all in the same column
  separate_rows(ninBeskrivelsesvariable, sep = ",") |>
  separate(
    col = ninBeskrivelsesvariable,
    into = c("NiN_variable_code", "NiN_variable_value"),
    sep = "_",
    remove = F
  ) |>
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
  filter(NiN_variable_code %in% myVars) |>
  select(
    id = identifikasjon_lokalId,
    municipality = kommunenummer,
    year = kartleggingsår,
    mosaic = mosaikk,
    quality = lokalitetskvalitet,
    biodiversity = naturmangfold,
    condition = tilstand,
    natureType = naturtype,
    variable = NiN_variable_code,
    value = NiN_variable_value,
    area
  ) |>
  st_transform(myCRS) # Choosing this to match the EDM (se further down)
# 19k obs.

naturetypes |>
  as_tibble() |>
  count(natureType, sort=T) |>
  mutate(natureType = fct_reorder(natureType, n)) |>
  ggplot(aes(x = natureType, y = n))+
  geom_col()+
  coord_flip()

naturetypes |>
  saveRDS("../data/surveyData.RDS")
# head(naturetypes)
```

```{r readSurvey}
#| eval: true

# read cached rds file. This is not so big.
naturetypes <- readRDS("../data/surveyData.RDS")
```

```{r convertToPercent}
#| eval: true

# I now want to take the variable and normalise it before I can then combine them
# despite them being on different scales
# I will first normalise by converting into %.
# Remeber the ordinal categories represnts frequency ranges

# The data is strongly right skewed, so simply taking the center value of each
# bin will not work.
naturetypes %>% 
  ggplot()+
  theme_bw() +
  geom_histogram(aes(x = value),
                 binwidth = 1) +
  facet_wrap(.~variable,
             scales = "free")

# I will use th lower bound for each bin.
# The exception in when the variable is 1, because then the lower bound
# is 0, same as when the variable is 0.
# For these I will set manually a slightly higher value.

naturetypes <- naturetypes %>%
  mutate(value = case_when(
    # selecting the variable that follow the same 4 step scale
    variable %in% c("7TK", "7SE", "7FA") ~ 
      case_match(value,
             0 ~ 0,
             1 ~ mean(c(0, 1/16))*100,
             2 ~ 1/16*100,
             3 ~ 50), # note that it is not possible to get a value of 1
    # selecting the eight step variables
    variable %in% c('PRTK', 'PRSL') ~ 
      case_match(value,
             0 ~ 0,
             1 ~ 1.5,
             2 ~ 3,
             3 ~ 6.25,
             4 ~ 12.5,
             5 ~ 25,
             6 ~ 50,
             7 ~ 75),
    .default = value
    )
  )

naturetypes %>% 
  ggplot()+
  theme_bw() +
  geom_histogram(aes(x = value),
                 binwidth = 1,
                 color = "orange",
                 fill = "orange") +
  facet_wrap(.~variable,
             scales = "free")

@@

```{r}
#| eval: true
# First I will combine 7TK and PRTK, and also 7SE and PRSL.

naturetypes_wide <- naturetypes_wide %>%
  mutate(TK = if_else(
    is.na(var_PRTK), var_7TK, var_PRTK),
    SE = if_else(
      is.na(var_PRSL), var_7SE, var_PRSL)
    )

plot_grid(
naturetypes_wide %>%
  as_tibble() |>
  count(SE,
        name = "sum") |>
  ggplot(
    aes(
      x = factor(SE),
      y = sum))+
  geom_bar(stat="identity",
           fill="grey",
           colour = "black")+
  theme_bw(base_size = 12)+
  labs(x = "7SE or PRSL score",
       y = "Number of localities"),

naturetypes_wide %>%
  as_tibble() |>
  count(TK,
        name = "sum") |>
  ggplot(
    aes(
      x = factor(TK),
      y = sum))+
  geom_bar(stat="identity",
           fill="grey",
           colour = "black")+
  theme_bw(base_size = 12)+
  labs(x = "7TK or PRTK score",
       y = "Number of localities")
)

# The NA's represnets localities where just one of the two variables 
# (then thinking 7SE and PRSL as the same variable)
# is recorded. To combine these into one metric ADSV, I could take the 
# one with the highest value (worst-rule) og the sum.
# Sum is problematic is not all locations ahve two values to sum together.
# But the other option is problematic since I think field workers often
# tend to split the effects over two variables is they have that option.
# And if we have 50% vehicle damaeg and 50% hiking damage, that is no doubt
# worst than just having 50% of either. So I will use the sum, despite its issues.
```

# Taking the sum of 7SE and 7TK (incl the PR.. variables)
naturetypes_wide <- naturetypes_wide |>
  rowwise() |>
  mutate(ADSV = sum(c(SE, TK), na.rm=TRUE))

naturetypes_wide %>%
  as_tibble() |>
  count(ADSV,
        name = "sum") |>
  ggplot(
    aes(
      x = (ADSV),
      y = sum))+
  geom_bar(stat="identity",
           fill="grey",
           colour = "black")+
  theme_bw(base_size = 12)+
  labs(x = "Summed ADVS score",
       y = "Number of localities") +
    scale_x_continuous(
    labels = scales::label_number(accuracy = 1))

@@

# Now I will copy these ADVS-values into the sf object again, keeping things in wide format
naturetypes <- naturetypes |>
  pivot_wider(#id_cols="id",
              names_from = "variable",
              values_from = "value") |>
  left_join(naturetypes_wide |> select(id, ADSV), by = "id") |>
  select(!c("7TK", "7SE", "PRSL", "PRTK")) 
  # >
  #pivot_longer(cols = c(ADSV, "7FA", "7GR-GI"),
  #             names_to = "variable",
  #             values_drop_na = T)
```

@@

# Adding scaled indicator values to the dataset
naturetypes$i_ADSV <- eaTools::ea_normalise(data = naturetypes,
                      vector = "ADSV",
                      upper_reference_level = lower,
                      lower_reference_level = upper,
                      break_point = threshold,
                      reverse = T
                      )

naturetypes$i_alien <- eaTools::ea_normalise(data = naturetypes,
                      vector = "7FA",
                      upper_reference_level = lower,
                      lower_reference_level = upper,
                      break_point = threshold,
                      reverse = T
                      )

naturetypes$i_ditch <- eaTools::ea_normalise(data = naturetypes,
                      vector = "7GR-GI",
                      upper_reference_level = lower2,
                      lower_reference_level = upper2,
                      break_point = threshold2,
                      reverse = T
                      )
```

```{r getNorwayOutline}
#| eval: true

outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
  st_transform(myCRS)
```

@@

# find the correct layer
# st_layers(path_muni)

# read inn data and transform
muni <- sf::read_sf(path_muni, layer = "kommune") |>
  st_transform(myCRS)

# This returns some multisurfaces as well
# table(st_geometry_type(muni))
# function to make sure that multipolygons are returned
ensure_multipolygons <- function(X) {
  tmp1 <- tempfile(fileext = ".gpkg")
  tmp2 <- tempfile(fileext = ".gpkg")
  st_write(X, tmp1)
  gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
  Y <- st_read(tmp2)
  st_sf(st_drop_geometry(X), geom = st_geometry(Y))
}

@@

# subset of the three target municipalities
muni3 <- muni |>
  filter(kommunenummer %in% c(
    "3020", # Nordre Follo
    "5001", # Trondheim
    "5418" # Målselv
  )) |>
  mutate(Municipality = case_when(
    kommunenummer == "3020" ~ "Nordre Follo",
    kommunenummer == "5001" ~ "Trondheim",
    kommunenummer == "5418" ~ "Målselv"
  ))

# To crop EDM, I need the three municipalities seprately.
nf <- muni3 |>
  filter(kommunenummer == "3020")
tr <- muni3 |>
  filter(kommunenummer == "5001")
me <- muni3 |>
  filter(kommunenummer == "5418")
```

@@

nature3 |>
  as_tibble() |>
  count(municipality,
    sort = TRUE,
    name = "Number of polygons")

# and also to get the data coverage polygon.
# These I had to manually download from naturbase.no
dk <- read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
  st_transform(st_crs(muni3)) |>
  st_intersection(muni3) |>
  st_union()
```

@@

(positionMap <- tm_shape(outline) +
  tm_polygons() +
  tm_shape(centroids) +
  tm_text(
    text = "Municipality",
    size = .8,
    xmod = 2,
    ymod = -1
  ) +
  tm_symbols(
    size = 1.5,
    shape = 17,
    col = "black"
  ) +
  tm_grid(projection = 4326) +
  tm_layout(
    bg.color = "skyblue",
    outer.margins = c(0.01, .02, .02, .02)))
```

>>>>

# Some directories

# Ecosystem delineation map
path_mire <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"

# infrastructure index
path_infrastructure <- ifelse(dir == "C:",
  "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif",
  "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif"
)

# field survey
# # downloaded from https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c
path_naturetypes <- "../data/survey.gdb"

# municipality outline
path_muni <- "../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb"

@@

```{r import}
#| cache: true

# I already did some work to identify the relevant nature types
# summary file (https://github.com/NINAnor/ecosystemCondition/blob/main/data/naturetypes/natureType_summary.rds)
naturetypes_summary <- readRDS("../data/natureType_summary.rds")

# Survey data
# The data is too big to be stored on GitHub
# Impart polygon data set
st_layers(path_naturetypes)
naturetypes <- sf::st_read(dsn = path_naturetypes, layer = "naturtyper_nin_omr")
# 142k polygons (2023)
# 
# Impart survey coverage map
coverage <- sf::st_read(dsn = path_naturetypes, layer = "naturtyper_nin_dekning") |>
  st_transform(myCRS)

# Outline of norway (coastline)
outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
  st_transform(myCRS)

# Municipalities
# find the correct layer
st_layers(path_muni)
# read inn data and transform
muni <- sf::read_sf(path_muni, layer = "kommune") |>
  st_transform(myCRS)

# Mire data
mire_terra <- terra::rast(path_mire)
```

@@

@@

```{r natureTypeData}
#| eval: true
#| cache: true

# Clean the survey data

naturetypes <- naturetypes |>
  # keep only wetlands
  filter(
    hovedøkosystem == "våtmark",
    naturtype %in% nts,
    naturtype != "Kalkrik helofyttsump"
  ) |>
  # calculate the areas (m2) of the polygons
  mutate(area = SHAPE |> st_area()) |>
  # the variable codes and values are all in the same column
  separate_rows(ninBeskrivelsesvariable, sep = ",") |>
  separate(
    col = ninBeskrivelsesvariable,
    into = c("NiN_variable_code", "NiN_variable_value"),
    sep = "_",
    remove = F
  ) |>
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
  filter(NiN_variable_code %in% myVars) |>
  select(
    id = identifikasjon_lokalId,
    municipality = kommunenummer,
    year = kartleggingsår,
    mosaic = mosaikk,
    quality = lokalitetskvalitet,
    biodiversity = naturmangfold,
    condition = tilstand,
    natureType = naturtype,
    variable = NiN_variable_code,
    value = NiN_variable_value,
    area
  ) |>
  st_transform(myCRS) # Choosing this to match the EDM (se further down)
# 19k obs.
```

```{r}
#| eval: false
# Plot to show what the most common nature types in the data set are
naturetypes |>
  as_tibble() |>
  count(natureType, sort=T) |>
  mutate(natureType = fct_reorder(natureType, n)) |>
  ggplot(aes(x = natureType, y = n))+
  geom_col()+
  coord_flip()
```

```{r convertToPercent}
#| eval: true
# I now want to take the variables and normalise them before I can then combine
# them despite them being on different scales.
# I will first normalise by converting into % (not for 7GR-GI).
# Remember the ordinal categories represents frequency ranges
# The data is strongly right skewed, so simply taking the center value of each
# bin will not work:
naturetypes %>%
  ggplot() +
  theme_bw() +
  geom_histogram(aes(x = value),
    binwidth = 1
  ) +
  facet_wrap(. ~ variable,
    scales = "free"
  )

# I will use the lower bound for each bin instead.
# The exception in when the variable is 1, because then the lower bound
# is 0, same as when the variable is 0.
# For these I will set manually a slightly higher value.

naturetypes <- naturetypes %>%
  mutate(value = case_when(
    # selecting the variables that follow the same 4 step scale
    variable %in% c("7TK", "7SE", "7FA") ~
      case_match(
        value,
        0 ~ 0,
        1 ~ mean(c(0, 1 / 16)) * 100,
        2 ~ 1 / 16 * 100,
        3 ~ 50
      ), # note that it is not possible to get a value of 1
    # selecting the eight step variables
    variable %in% c("PRTK", "PRSL") ~
      case_match(
        value,
        0 ~ 0,
        1 ~ 1.5,
        2 ~ 3,
        3 ~ 6.25,
        4 ~ 12.5,
        5 ~ 25,
        6 ~ 50,
        7 ~ 75
      ),
    .default = value
  ))

naturetypes %>%
  filter(variable != "7GR-GI") |>
  ggplot() +
  theme_bw() +
  geom_histogram(aes(x = value),
    binwidth = 1,
    color = "orange",
    fill = "orange"
  ) +
  xlab("%") +
  facet_wrap(. ~ variable,
    scales = "free"
  )

@@

```{r}
#| eval: true
# First I will combine 7TK and PRTK, and also 7SE and PRSL.
naturetypes_wide <- naturetypes_wide %>%
  mutate(
    TK = if_else(
      is.na(var_PRTK), var_7TK, var_PRTK
    ),
    SE = if_else(
      is.na(var_PRSL), var_7SE, var_PRSL
    )
  )

plot_grid(
  naturetypes_wide %>%
    as_tibble() |>
    count(SE,
      name = "sum"
    ) |>
    ggplot(
      aes(
        x = factor(SE),
        y = sum
      )
    ) +
    geom_bar(
      stat = "identity",
      fill = "grey",
      colour = "black"
    ) +
    theme_bw(base_size = 12) +
    labs(
      x = "7SE or PRSL score",
      y = "Number of localities"
    ),
  naturetypes_wide %>%
    as_tibble() |>
    count(TK,
      name = "sum"
    ) |>
    ggplot(
      aes(
        x = factor(TK),
        y = sum
      )
    ) +
    geom_bar(
      stat = "identity",
      fill = "grey",
      colour = "black"
    ) +
    theme_bw(base_size = 12) +
    labs(
      x = "7TK or PRTK score",
      y = "Number of localities"
    )
)

# The NA's represents localities where just one of the two variables
# (then thinking 7SE and PRSL as the same variable)
# is recorded. To combine these into one metric ADSV, I could take the
# one with the highest value (worst-rule) og the sum.
# Sum is problematic is not all locations have two values to sum together.
# But the other option is problematic since I think field workers often
# tend to split the effects over two variables is they have that option.
# And if we have 50% vehicle damage and 50% hiking damage, that is no doubt
# worst than just having 50% of either. So I will use the sum, despite its issues.
```

```{r}
#| eval: true
# Taking the sum of 7SE and 7TK (incl the PR.. variables)
naturetypes_wide <- naturetypes_wide |>
  rowwise() |>
  mutate(ADSV = sum(c(SE, TK), na.rm = TRUE))

naturetypes_wide %>%
  as_tibble() |>
  count(ADSV,
    name = "sum"
  ) |>
  ggplot(
    aes(
      x = ADSV,
      y = sum
    )
  ) +
  geom_bar(
    stat = "identity",
    fill = "grey",
    colour = "black"
  ) +
  theme_bw(base_size = 12) +
  labs(
    x = "Summed ADVS score",
    y = "Number of localities"
  ) +
  scale_x_continuous(
    labels = scales::label_number(accuracy = 1)
  )

@@

# Now I will copy these ADVS-values into the sf object again, keeping things in wide format
naturetypes <- naturetypes |>
  pivot_wider(
    names_from = "variable",
    values_from = "value"
  ) |>
  left_join(naturetypes_wide |> select(id, ADSV), by = "id") |>
  select(!c("7TK", "7SE", "PRSL", "PRTK"))
  # >
  #pivot_longer(cols = c(ADSV, "7FA", "7GR-GI"),
  #             names_to = "variable",
  #             values_drop_na = T)
head(naturetypes)
```

@@

```{r}
#| eval: true
# Adding scaled indicator values to the dataset
naturetypes$i_ADSV <- eaTools::ea_normalise(
  data = naturetypes,
  vector = "ADSV",
  upper_reference_level = lower,
  lower_reference_level = upper,
  break_point = threshold,
  reverse = T
)

naturetypes$i_alien <- eaTools::ea_normalise(
  data = naturetypes,
  vector = "7FA",
  upper_reference_level = lower,
  lower_reference_level = upper,
  break_point = threshold,
  reverse = T
)

naturetypes$i_ditch <- eaTools::ea_normalise(
  data = naturetypes,
  vector = "7GR-GI",
  upper_reference_level = lower2,
  lower_reference_level = upper2,
  break_point = threshold2,
  reverse = T
)
```

@@

# The data contains some multisurfaces as well
# table(st_geometry_type(muni))
# function to make sure that multipolygons are returned
ensure_multipolygons <- function(X) {
  tmp1 <- tempfile(fileext = ".gpkg")
  tmp2 <- tempfile(fileext = ".gpkg")
  st_write(X, tmp1)
  gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
  Y <- st_read(tmp2)
  st_sf(st_drop_geometry(X), geom = st_geometry(Y))
}

@@

# subset of the three target municipalities
muni3 <- muni |>
  filter(kommunenummer %in% c(
    "3020", # Nordre Follo
    "3451", # Nord-Aurdal
    "3446" # Gran
  )) |>
  mutate(Municipality = case_when(
    kommunenummer == "3020" ~ "Nordre Follo",
    kommunenummer == "3451" ~ "Nord-Aurdal",
    kommunenummer == "3446" ~ "Gran"
  ))

# To crop EDM, I need the three municipalities seprately.
nf <- muni3 |>
  filter(kommunenummer == "3020")
na <- muni3 |>
  filter(kommunenummer == "3451")
gr <- muni3 |>
  filter(kommunenummer == "3446")
```

@@

nature3 |>
  as_tibble() |>
  count(municipality,
    sort = TRUE,
    name = "Number of polygons")
# There where some polygons that spanned municipal borders. 
# It's not a problem

# and also to get the data coverage polygon.
coverage3 <- coverage |>
  st_intersection(muni3)

## These I had to manually download from naturbase.no
#dk <- read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
#  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
#  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
#  st_transform(st_crs(muni3)) |>
#  st_intersection(muni3) |>
#  st_union()
```

@@

inc <- 1e5
myBbox <- st_bbox(centroids)
myBbox[1:2] <- myBbox[1:2]-inc 
myBbox[3:4] <- myBbox[3:4]+inc

(positionMap <- 
  tm_shape(outline,
           bbox = myBbox) +
  tm_polygons() +
  tm_shape(centroids) +
  tm_text(
    text = "Municipality",
    just= "left",
    size = .8,
    xmod = 1,
    ymod = 0
  ) +
  tm_symbols(
    size = 1.5,
    shape = 17,
    col = "black"
  ) +
  tm_grid(projection = 4326) +
  tm_layout(
    bg.color = "skyblue",
    outer.margins = c(0.01, .02, .02, .02))+
  tm_compass()+
  tm_scale_bar()+
    tm_basemap(leaflet::providers$Stamen.Watercolor)
)
